---
title: "Annotation of BM+CB data across multiple datasets with Conos"
output: rmarkdown::github_document
vignette: >
  %\VignetteIndexEntry{Annotation of BM+CB data across multiple datasets with Conos}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette shows annotation of BM+CB dataset from the [Conos tutorial](https://github.com/hms-dbmi/conos/blob/master/vignettes/walkthrough.md) across multiple samples

```{r setup, message=FALSE, warning=FALSE}
library(CellAnnotatoR)
library(conos)
library(pagoda2)
library(dplyr)
library(ggplot2)
library(pbapply)

theme_set(theme_bw())
```

## Pre-processing

Let's load and pre-process data:

```{r, message=FALSE, warning=FALSE}
panel <- readRDS(file.path(find.package('conos'),'extdata','panel.rds'))
panel_preprocessed <- lapply(panel, basicP2proc, n.cores=4, min.cells.per.gene=0, 
                             n.odgenes=2e3, get.largevis=FALSE, make.geneknn=FALSE)
```

```{r}
cellannot <- read.table(file.path(find.package('conos'),'extdata','cellannot.txt'),header=F,sep='\t')
cellannot <- setNames(cellannot[,2], cellannot[,1])
```

Now we can integrate it with Conos:

```{r}
con <- Conos$new(panel_preprocessed, n.cores=4)
con$buildGraph()
con$findCommunities(method=conos::leiden.community, resolution=5)
con$embedGraph(method="UMAP", min.dist=1, spread=2, n.cores=30)

con$plotGraph(size=0.2)
```

Prepare data for annotation:

```{r}
marker_path <- rprojroot::find_package_root_file("markers/bm_cb.md")
clf_datas <- pblapply(con$samples, function(p2) 
  getClassificationData(Matrix::t(p2$misc$rawCounts), marker_path))

marker_scores <- pblapply(clf_datas, getMarkerScoresPerCellType)
```

## Annotation

Now we can run individual annotation on each dataset:

```{r}
ann_by_dataset <- pbmapply(function(cd, ms, p2) 
  assignCellsByScores(p2$graphs$PCA, scores=ms, clf.data=cd),
  clf_datas, marker_scores, panel_preprocessed, SIMPLIFY=F) %>% 
  setNames(names(clf_datas))

all_annotations <- lapply(ann_by_dataset, function(an) an$annotation$l1) %>% Reduce(c, .)
con$plotGraph(groups=all_annotations, size=0.2, plot.na=F)
```

We can see that running annotation on individual samples doesn't neccesserily guarantee smoothness of labeling on the joint graph, 
as such approach can't utilize joint structure. To deal with it we can run annotation on the whole graph:

```{r}
all_scores <- Reduce(rbind, marker_scores)
ann_by_level <- assignCellsByScores(con$graph, scores=all_scores, clf.data=clf_datas[[1]], score.fixing.threshold=0.99)
con$plotGraph(groups=ann_by_level$annotation$l1, size=0.2, plot.na=F)
```

To further deal with noise, we can use clustering information:

```{r}
clusters <- con$clusters$leiden$groups
annotation <- assignCellsByScores(con$graph, scores=all_scores, clf.data=clf_datas[[1]], 
                                  clusters=clusters)$annotation$l1
con$plotGraph(groups=annotation, size=0.2)
# con$plotGraph(groups=clusters, size=0.2, shuffle.colors=T)
```

In the current example, clustering resolution is too low to separate all subpopulatons, which lead to lack of CLP and DC populations. Let's increase resolution:

```{r}
target_clusters <- clusters[names(annotation)[annotation %in% c("Progenitors", "Plasma")]] %>% 
  as.character() %>% unique()
clusters_inc <- findSubcommunities(con, target_clusters, groups=clusters)
con$plotGraph(groups=clusters_inc, size=0.2, shuffle.colors=T)
```

And now we can re-run annotation:

```{r}
annotation <- assignCellsByScores(con$graph, scores=all_scores, clf.data=clf_datas[[1]], 
                                  clusters=clusters_inc)$annotation$l1
con$plotGraph(groups=annotation, size=0.2)
```
