---
title: "Conos alignment of new datasets"
output: rmarkdown::github_document
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

```{r, message=FALSE, warning=FALSE}
library(ggplot2)
library(magrittr)
library(Matrix)
library(pbapply)
library(dplyr)
library(pagoda2)
library(CellAnnotatoR)

theme_set(theme_bw())
```

## Download data

To run this vignette you need [MCA](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE108097) Lung data, which is available under accession numbers 
[GSM2906429](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM2906429), [GSM2906430](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM2906430) 
and [GSM2906431](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM2906431). There, you need to download files *GSM2906429_Lung1_dge.txt.gz*, *GSM2906430_Lung2_dge.txt.gz* and *GSM2906431_Lung3_dge.txt.gz*. 

```
wget https://www.ncbi.nlm.nih.gov/geo/download/\?acc\=GSM2906429\&format\=file\&file\=GSM2906429%5FLung1%5Fdge%2Etxt%2Egz -O ./GSM2906429_Lung1_dge.txt.gz
wget https://www.ncbi.nlm.nih.gov/geo/download/\?acc\=GSM2906430\&format\=file\&file\=GSM2906430%5FLung2%5Fdge%2Etxt%2Egz -O ./GSM2906430_Lung2_dge.txt.gz
wget https://www.ncbi.nlm.nih.gov/geo/download/\?acc\=GSM2906431\&format\=file\&file\=GSM2906431%5FLung3%5Fdge%2Etxt%2Egz -O ./GSM2906431_Lung3_dge.txt.gz
```

You also need to download `MCA_CellAssignments.csv` from the [paper figshare](https://figshare.com/s/865e694ad06d5857db4b):

```
wget https://ndownloader.figshare.com/files/11083451\?private_link\=865e694ad06d5857db4b -O MCA_CellAssignments.csv
```

Then, extract the content with `gunzip GSM2906429_Lung1_dge.txt.gz GSM2906430_Lung2_dge.txt.gz GSM2906431_Lung3_dge.txt.gz`.

## Load data

Here, function `dataPath` contains path to the downloaded files with data. You need to adjust it to your folder:

```{r}
dataPath <- function(...) file.path("~/mh/Data/MCA/lung/", ...)

readTxtMtx <- function(path) {
  suppressWarnings(cm <- data.table::fread(path))
  cm <- as.matrix(cm[, 2:ncol(cm)], rownames=cm$V1) %>% as("dgCMatrix")
  return(cm)
}
```

```{r}
cms <- list.files(dataPath(), pattern="*.txt", full.names=T) %>% lapply(readTxtMtx)

cm_merged <- conos:::mergeCountMatrices(cms)
cell_info <- read.csv(dataPath("MCA_CellAssignments.csv"))

cluster_ids <- cell_info %$% setNames(gsub("Lung_", "", ClusterID), Cell.name)
annotation <- cell_info %$% setNames(Annotation, Cell.name) %>% .[colnames(cm_merged)] %>% 
  droplevels() %>% gsub("[(]Lung[)]", "", .) %>% gsub("−", "-", .) %>% 
  setNames(colnames(cm_merged)) %>% .[!is.na(.)]
```

Here we use Pagoda2 to get embedding and differential genes. In case your data consists of multiple datasets, Conos or Seurat may be used as well.

```{r, fig.width=5, fig.height=5}
p2 <- basicP2proc(cm_merged[, names(annotation)], n.cores=10, get.largevis=F, make.geneknn=F)
conos::embeddingPlot(p2$embeddings$PCA$tSNE, groups=annotation, font.size=c(2, 3))
```

## Define hierarchy

Hierarchies can be defiened either by hands or in an automated way. 

### Automated inference of hierarchies

The automated algorithm is rather simple and will be improved in future.  And currently, it **can not** be used on the aligned data.

```{r, fig.width=7, fig.height=8}
hierarchy <- deriveHierarchy(p2$reductions$PCA, annotation, max.depth=3)
clf_tree <- hierarchyToClassificationTree(hierarchy)
plotTypeHierarchy(clf_tree)
```

We can give meaningful names to the most of the derived types. As Pagoda PCA is stochastic, hierarchies may 
slightly vary from run to run, so these lines must be adjusted for your cases:

```{r, eval=FALSE}
t_names <- c(l1_1="Alveolar1", l1_4="B", l1_6="Ciliated", l2_14="Endothelial1", l1_10="Endothelial",
             l2_21="Stromal1", l1_11="Stromal", l2_15="Granulocyte", l2_12="T", l1_9="T_NK", 
             l2_2="Macrophage1", l2_2="Macrophage", l2_8="Dendritic1", l2_10="Dendritic2", l1_7="Dendritic0")

igraph::V(clf_tree)[names(t_names)]$name <- t_names
```

### Manual definition of hierarchies in code

Alternatively, you can define hierarchy manually:

```{r, fig.width=7, fig.height=8}
hierarchy <- list(
  Alveolar=c("Alveolar bipotent progenitor", "AT1 Cell", "AT2 Cell"),
  B=c("B Cell", "Ig-producing B cell"),
  Dendritic=list(Dendritic1=c("Conventional dendritic cell_Tubb5 high", "Dendritic cell_Naaa high", "Dividing dendritic cells", 
                              "Conventional dendritic cell_H2-M2 high", "Conventional dendritic cell_Mgl2 high"), 
                 Dendritic2=c("Conventional dendritic cell_Gngt2 high", "Plasmacytoid dendritic cell")),
  Endothelial=c("Endothelial cell_Kdr high", "Endothelial cell_Tmem100 high", "Endothelial cells_Vwf high"),
  Granulocyte=c("Eosinophil granulocyte", "Neutrophil granulocyte"),
  Macrophage=c("Alveolar macrophage_Ear2 high", "Alveolar macrophage_Pclaf high", "Interstitial macrophage"),
  Stromal=c("Stromal cell_Acta2 high", "Stromal cell_Dcn high", "Stromal cell_Inmt high"),
  NK_T=list(`T`=c("Dividing T cells", "T Cell_Cd8b1 high", "Nuocyte"), "NK Cell")
) %>% c(setdiff(unique(as.character(annotation)), unlist(.)))

clf_tree <- hierarchyToClassificationTree(hierarchy)
plotTypeHierarchy(clf_tree)
```

## Marker selection

Original annotation by level:

```{r, fig.width=10, fig.height=4}
ann_by_level <- mergeAnnotationByLevels(annotation, clf_tree)
plotAnnotationByLevels(p2$embeddings$PCA$tSNE, ann_by_level, font.size=c(2, 3), n.col=3, shuffle.colors=T)
```

First, we need to find DE genes for each sub-brunch of the hierarchy:

```{r}
ann_by_parent <- getAnnotationPerParent(clf_tree, annotation)

de_info_per_parent <- ann_by_parent %>% 
  pblapply(function(ann) p2$getDifferentialGenes(groups=ann, z.threshold=0))

de_info_per_parent <- pbmapply(function(de, ann) prepareDeInfo(de, ann, cm.raw=p2$misc$rawCounts, n.cores=10), 
                               de_info_per_parent, ann_by_parent)
```

Next, do TF-IDF normalization of the matrix:

```{r}
cm_norm <- cm_merged %>% normalizeTfIdfWithFeatures() %>% .[names(annotation), ]
```

Now we can define blacklist of the genes, which shouldn't be used as markers:

```{r}
marker_blacklist <- rownames(cm_merged) %>% .[grep("mt-", .)]
```

Finally, we run marker selection for each of the sub-branches:

```{r}
marker_info <- names(de_info_per_parent) %>% setNames(., .) %>% lapply(function(n)
  selectMarkersPerType(cm_norm, ann_by_parent[[n]], 
                       preSelectMarkerCandidates(de_info_per_parent[[n]], blacklist=marker_blacklist), 
                       parent=n, max.iters=50, max.uncertainty=0.25, log.step=5, verbose=1, n.cores=10))

marker_list <- setNames(marker_info, NULL) %>% unlist(recursive=F)
```

## Improving known list of markers

If you already know some markers, this knowledge can be used. You can either read the markers from file or create marker list by hands in code.

Reading markers:

```{r, eval=FALSE}
marker_list_prior <- parseMarkerFile("markers_prior.txt")
```

Create by hands:

```{r}
marker_list_prior <- emptyMarkerList(clf.tree=clf_tree)
marker_list_prior$`Conventional dendritic cell_Mgl2 high`$expressed <- c("Mgl2")
marker_list_prior$`Dendritic cell_Naaa high`$expressed <- c("Naaa")
```

Check, which markers aren't presented in the matrix:

```{r}
lapply(marker_list_prior, `[`, c("expressed", "not_expressed")) %>% unlist() %>% 
  setdiff(colnames(cm_norm))
```

```{r}
marker_info <- names(de_info_per_parent) %>% setNames(., .) %>% lapply(function(n)
  selectMarkersPerType(cm_norm, ann_by_parent[[n]], 
                       preSelectMarkerCandidates(de_info_per_parent[[n]], blacklist=marker_blacklist), 
                       marker.list=marker_list_prior,
                       parent=n,max.iters=50, max.uncertainty=0.25, log.step=5, verbose=1, n.cores=10))

marker_list <- setNames(marker_info, NULL) %>% unlist(recursive=F)
```

## Diagnostics

Number of markers per type:

```{r}
sapply(marker_list, sapply, length)[1:2,] %>% t()
```

To evaluate quality of these markers we can annotate cell using them and use the uncertainty plots.

```{r}
clf_data <- getClassificationData(cm_norm, marker_list, prenormalized=T)
ann_info <- assignCellsByScores(NULL, clf_data, clusters=annotation)
```

Original annotation:

```{r, fig.width=10, fig.height=4}
plotAnnotationByLevels(p2$embeddings$PCA$tSNE, ann_by_level, n.col=3, font.size=c(2, 3))
```

New annotation:

```{r, fig.width=10, fig.height=4}
plotAnnotationByLevels(p2$embeddings$PCA$tSNE, ann_info$annotation, n.col=3, font.size=c(2, 3))
```

Uncertainty plots per cell:

```{r, fig.width=8, fig.height=6}
score_info <- getMarkerScoreInfo(clf_data)
unc_per_cell <- scoreCellUncertaintyPerLevel(ann_info, score_info)

lapply(unc_per_cell, function(unc) plotUncertaintyPerCell(p2$embeddings$PCA$tSNE, unc)) %>% 
  cowplot::plot_grid(plotlist=., ncol=1, labels=names(unc_per_cell), label_y=0.9, label_x=0.01)
```

Level 1 assignment scores shown on cell embedding:

```{r, fig.width=10, fig.height=6}
plotAssignmentScores(p2$embeddings$PCA$tSNE, ann_info$scores$l1, clf_data$classification.tree, parent="root", build.panel=T, size=0.1, n.col=5)
```

Positive markers:

```{r, fig.height=10, fig.width=7}
marker_list[sort(unique(ann_by_level$l1))] %>% lapply(`[[`, "expressed") %>% 
  unlist() %>% unique() %>% 
  plotExpressionViolinMap(p2$counts, ann_by_level$l1, gene.order=T)
```

## Save to a file

```{r}
markerListToMarkup(marker_list, file="mca_markers_auto.txt")
```
